---
title: "Lesson 1: Visualizing data with *ggplot2*"
date: today
format:
    html:
        standalone: true
        embed-resources: true
        number-sections: true
        toc: true
editor: visual
---

# Prepare the R envinronment for this lesson

When we start R, we only have access to the **base R** packages. In order to use
any additional packages, we need to load them into memory with the `library()`
function.

For this lesson we'll need two packages: *ggplot2* and *palmerpenguins*. Run the
following code chunk to load these two libraries. Note, the second library
command is empty. Modify the command so it loads the palmerpenguins package.

```{r}
library(ggplot2)
library(palmerpenguins)
```

It's good practice to start R code, stored in R scripts or quarto/Rmarkdown
documents, by loading the packages we'll be using in the body of the code. It
gives readers, including future versions of ourselves, a consistent place to
examine all of the packages required to run our code.




```{r}
#| label: Completed flipper length vs body mass graph
#| fig-width: 7
#| fig-height: 5
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
    geom_point(aes(color = species, shape = species)) +
    geom_smooth(method = "lm") +
    labs(title = "Relationship between flipper length and body mass",
         subtitle = "Across three species of penguins studied at the Palmer Research Station",
         x = "Body mass (g)",
         y = "Flipper length (mm)") +
    # scale_color_viridis_d() +
    # scale_color_brewer(palette = "Dark2") +
    theme_minimal()
```




# R session information

For reproducibility, we report version numbers for R and the packages we used in
our analyses above.

```{r}
sessionInfo()
```
